BIP-Predictor

### 2025-05-08 Latex/BIP_predictor.tex
up

% latex table generated in R 4.2.2 by xtable 1.8-4 package
% Thu May  8 07:34:02 2025
\begin{table}[ht]
\centering
\begin{tabular}{rrrrrr}
  \hline
 & GDP & ip & ifo\_c & ESI & spr\_10y\_3m \\ 
  \hline
2024-08-01 &  & 93.200 & 84.900 & 90.900 & -1.400 \\ 
  2024-09-01 &  & 91.200 & 83.700 & 89.800 & -1.400 \\ 
  2024-10-01 & 900.761 & 90.800 & 84.100 & 90.700 & -1.200 \\ 
  2024-11-01 &  & 92.200 & 83.800 & 89.300 & -0.900 \\ 
  2024-12-01 &  &  & 82.800 & 86.900 & -0.800 \\ 
  2025-01-01 &  &  & 82.500 & 88.100 & -0.200 \\ 
   \hline
\end{tabular}
\caption{Ragged end.  } 
\label{f_stat}
\end{table}% latex table generated in R 4.2.2 by xtable 1.8-4 package
% Thu May  8 07:34:02 2025
\begin{table}[ht]
\centering
\begin{tabular}{rrrrrr}
  \hline
 & GDP & ip & ifo\_c & ESI & spr\_10y\_3m \\ 
  \hline
2024-05-01 &  & 93.400 & 84.900 & 90.900 & -1.400 \\ 
  2024-06-01 &  & 90.700 & 83.700 & 89.800 & -1.400 \\ 
  2024-07-01 & 902.571 & 93.200 & 84.100 & 90.700 & -1.200 \\ 
  2024-08-01 &  & 91.200 & 83.800 & 89.300 & -0.900 \\ 
  2024-09-01 &  & 90.800 & 82.800 & 86.900 & -0.800 \\ 
  2024-10-01 & 900.761 & 92.200 & 82.500 & 88.100 & -0.200 \\ 
   \hline
\end{tabular}
\caption{Data aligned at sample end.} 
\label{f_stat}
\end{table}

@@

So-called direct forecasts are obtained by regressing the relevant indicators on forward-shifted BIP, accounting for the additional publication lag. For illustration of the concept, we here rely on all indicators and full sample information, excluding the pandemic. In-sample forecasts and the shifted BIP, which serves as the target, are presented in Figs. \ref{direct_wc} (full dataset without the pandemic) and \ref{direct_wc_financial_crisis} (financial crisis) for horizons up to three quarters ahead. 
\begin{figure}[H]\begin{center}\includegraphics[height=4in, width=5in]{./Figures/direct_wc_all.pdf}\caption{Direct forecasts (red lines) based on all indicators and full sample information, without the pandemic: BIP (black line) shifted by zero one, two and three quarters.\label{direct_wc}}\end{center}\end{figure}\begin{figure}[H]\begin{center}\includegraphics[height=4in, width=5in]{./Figures/direct_wc_financial_crisis.pdf}\caption{Direct forecast  across the financial crisis:  regression based on all indicators and full sample information.\label{direct_wc_financial_crisis}}\end{center}\end{figure}The results indicate that as $h$ increases, the peaks and dips of the direct forecasts become increasingly right-shifted (delayed) relative to the target. Additionally, the F-statistics precipitously decline from $F=3.57$, for $h=2$, to $F=0.46$, for $h=3$, see Table \eqref{f_stat}, suggesting that the predictor becomes statistically insignificant for larger forecast horizons exceeding two quarters (link to Heinisch and Scheufele).

@@

As shown, the performance of direct forecasts sharply declines for larger forecast horizons. We conjecture that erratic short-term fluctuations -—unpredictable high-frequency noise -— overlay the data, thereby obscuring the effective 'signal' and making a direct regression more susceptible to overfitting. We therefore envisage to highlight the signal by applying a filter to dampen high-frequency noise. 
The so-called Hodrick-Prescott filter is a classic tool used in business cycle analysis: the filter is specified by a single smoothing parameter $\lambda$ and the value $\lambda=1600$ is specifically recommended for quarterly data. However, Phillips and Jin (2021) suggest that the HP(1600) filter removes relevant information due to excessive smoothing.  In our context, this oversmoothing issue would be further exacerbated when considering forecast horizons shorter than a year -—an interval inconsistent with the mean duration of up to several years of business cycles, as highlighted by the HP(1600). Consequently, we here select a more adaptive HP(160) 'target' filter. 
While this choice may appear somewhat arbitrary, an analysis of the filter in the frequency domain confirms a more suitable profile for the corresponding amplitude function, assigning greater weight to yearly components -—relevant in our prediction framework-— than the classic quarterly HP filter. This approach also addresses and controls for undesirable high-frequency noise. See Fig. \ref{hp_160}, which presents both two-sided and (classic) one-sided filters (top panels) along with the associated amplitude functions (bottom panels). We now posit the HP(160) filter as a means to achieve better forecasting performance at larger horizons, noting that a comprehensive technical analysis of the effect of $\lambda$ on the resulting BIP predictor can be conducted using the M-SSA package. 
\begin{figure}[H]\begin{center}\includegraphics[height=3in, width=4.5in]{./Figures/hp_160.pdf}\caption{Two-sided HP-160) (top left) and one-sided (concurrent) HP-C filters (top right) with corresponding amplitude functions (bottom).\label{hp_160}}\end{center}\end{figure}
\subsection{Direct HP Forecast}\label{hpdf}

Extending the `classic' direct forecast design, we here consider filtered indicators, based on the one-sided HP(160), denoted as HP-C: the filtered indicators are explanatory variables for the regressions on forward-shifted BIP\footnote{Katja and Simon: for each forecast horizon $h=0,1,2,3$ I'm using corresponding forecasts of the real-time filter output, assuming the data to be white noise (assuming an ARMA model does not lead to significantly different results). I think we might ignore these technical details in the paper.}. As in the previous section, we rely on all indicators as well as the full sample estimate for illustration, see Figs.\eqref{direct_hp_forecasts} (all observations with exclusion of the pandemic) and \eqref{direct_hp_forecasts_financial_crisis} (financial crisis). A comparison with the classic direct forecasts in Fig.\eqref{direct_wc_financial_crisis} indicates that the new predictors are slightly less retarded (`faster'). 
\begin{figure}[H]\begin{center}\includegraphics[height=4in, width=5in]{./Figures/direct_hp_forecasts.pdf}\caption{Direct HP forecasts: entire data set with exclusion of the pandemic.\label{direct_hp_forecasts}}\end{center}\end{figure}\begin{figure}[H]\begin{center}\includegraphics[height=5in, width=6in]{./Figures/direct_hp_forecasts_financial_crisis.pdf}\caption{Shifted GDP (black), direct HP forecast (blue) and classic direct forecast (red) over the curse of the financial crisis:  regression based on all indicators and full sample information (without pandemic).\label{direct_hp_forecasts_financial_crisis}}\end{center}\end{figure}Moreover, the F-statistics are larger than for the direct forecasts, see Table \eqref{f_stat}, but the predictor remains statistically insignificant at forecast horizons exceeding two quarters (even though the evaluation is conducted in-sample and based on all available indicators). 
% latex table generated in R 4.2.2 by xtable 1.8-4 package
% Thu May  8 07:34:04 2025
\begin{table}[ht]
\centering
\begin{tabular}{rrrrr}
  \hline
 & h=0 & h=1 & h=2 & h=3 \\ 
  \hline
Direct forecast & 7.188 & 5.285 & 3.574 & 0.460 \\ 
  Direct HP forecast & 21.886 & 8.415 & 4.810 & 1.834 \\ 
   \hline
\end{tabular}
\caption{F-statistics of direct forecasts and direct HP forecasts for forecast horizons $h=0,1,2,3$: fully in-sample.  } 
\label{f_stat}
\end{table}We now propose a more refined causal filter design aiming for statistical significance of the resulting predictor at longer forecast horizons, out-of-sample.

@@

Figure \ref{bk_gammak} displays the coefficients of the multivariate nowcasts, where increased smoothing achieved through the M-SSA method manifests as a distinct pattern in the filter weights, notably influencing the rate of decay across increasing lags. The resulting filtered series, presented in Figure \ref{mssa_msse_now}, are standardized to enable direct comparisons. The M-SSA nowcast adheres to the specified constraint, exhibiting approximately one-third fewer zero-crossings than the M-MSE, confirming the effectiveness of the smoothness control imposed during the design. This is further corroborated by Table \ref{corhtnow}, which reports empirical holding times and target correlations. Increased smoothness achieved by the M-SSA comes at the cost of a slight reduction in tracking accuracy, as indicated by a marginally lower target correlation. This highlights the inherent trade-off—often referred to as the prediction dilemma—between smoothness (zero-crossing rate) and predictive accuracy (target correlation) within the M-SSA criterion.
\begin{figure}[H]\begin{center}\includegraphics[height=3in, width=4.5in]{./Figures/bk_gammak.pdf}\caption{Multivariate M-MSE (left) and M-SSA  (right) nowcast filters.\label{bk_gammak}}\end{center}\end{figure}\begin{figure}[H]\begin{center}\includegraphics[height=3in, width=4.5in]{./Figures/mssa_msse_now.pdf}\caption{Two-sided HP(160) applied to BIP (target, black) and nowcasts:  M-SSA (blue) and  M-MSE (green). The dashed vertical line delimits in- and out-of-sample spans. The first L=31 observations are lost due to initialization and the two-sided filter does not extend to the sample end.\label{mssa_msse_now}}\end{center}\end{figure}% latex table generated in R 4.2.2 by xtable 1.8-4 package
% Thu May  8 07:34:04 2025
\begin{table}[ht]
\centering
\begin{tabular}{rrr}
  \hline
 & M-SSA & M-MSE \\ 
  \hline
Target correlation & 0.68 & 0.69 \\ 
  HT & 6.50 & 3.71 \\ 
   \hline
\end{tabular}
\caption{Sample target correlations and HTs of M-SSA and M-MSE nowcasts.  } 
\label{corhtnow}
\end{table}The results validate that the smoothing constraint incorporated into the multivariate extension of Criterion \eqref{critssa} effectively manages the zero-crossings of the predictor: sample estimates of the holding times (HTs) align with the expected values derived from Equation \eqref{ht}\footnote{Effective convergence of sample estimates to their theoretical expected values —assuming the true data-generating process is known— can be validated using the M-SSA package. This validation is typically performed through analyses of very long samples of simulated (multivariate) data, ensuring the consistency of the estimation procedures under idealized conditions.}. An increase of $50\%$ in the HT relative to the M-MSE predictor has only a modest effect on the target correlation, suggesting that this tradeoff could be beneficial in our application. Going forward, we now omit the M-MSE, as it is a special case of the more general M-SSA framework.

@@

M-SSA predictors are compared to the traditional univariate HP-C in Figs. \eqref{multivar_vs_univar} and \eqref{mssa_hpc_financial_crisis}. In the first figure, we compare both filters across the periods of the financial crisis (top panels) and the COVID-19 pandemic (bottom panels), for forecast horizons $h=0,...,6$. The nowcast from the previous section is included for reference. In the second figure we compare both filters to the forward-shifted BIP for shifts larger than two quarters. All series are standardized to facilitate visual comparison. The primary distinction of the M-SSA approach is the progressive left-shift of the predictor as a function of $h$; this shift becomes more pronounced with increasing $h$ and appears to be consistent across all levels, including the peaks and troughs of the series. Conversely, the positions of peaks and troughs in the HP-C series appear to remain largely unaffected by variations in $h$. The more pronounced and systematic left-shift observed in the M-SSA forecasts can be partly attributed to the increasing weight assigned to the additional indicators leading BIP, which dominate the BIP component at larger forecast horizons. %In comparison, the effect of phase shifts—arising from the cyclical nature of the HP-filter and manifesting in the HP-C forecasts (shown in the left panels)—seems less systematic. 
\begin{figure}[H]\begin{center}\includegraphics[height=3in, width=4.5in]{./Figures/bk_h.pdf}\caption{Multivariate M-SSA: $h=0$ (nowcast, left) and $h=4$ (one year forecast, right).\label{bk_h}}\end{center}\end{figure}\begin{figure}[H]\begin{center}\includegraphics[height=3in, width=4.5in]{./Figures/multivar_vs_univar.pdf}\caption{Now- and forecasts of HP-BIP for $h=0,...,6$: HP-C (top) vs. M-SSA (bottom) across the financial crisis (left) and the pandemic (right).\label{multivar_vs_univar}}\end{center}\end{figure}
\begin{figure}[H]\begin{center}\includegraphics[height=3in, width=4.5in]{./Figures/mssa_hpc_financial_crisis.pdf}\caption{Forward-shifted BIP (black), M-SSA (blue) and HP-C (red) for $h=3,4,5,6$ across the financial crisis. All series are standardized for better visual inspection.\label{mssa_hpc_financial_crisis}}\end{center}\end{figure}

@@

There are various combinations of M-SSA components for tracking BIP. For illustration, we here select the single BIP component resulting from the M-SSA filter tracking HP-BIP. Table \eqref{p_val_hpbip_wc} reports HAC-adjusted p-values from regressions of the out-of-sample predictor on HP-BIP, indicating a strong link between the predictor and the target, out-of-sample. 
% latex table generated in R 4.2.2 by xtable 1.8-4 package
% Thu May  8 07:34:09 2025
\begin{table}[ht]
\centering
\begin{tabular}{rrrrrrrr}
  \hline
 & h=0 & h=1 & h=2 & h=3 & h=4 & h=5 & h=6 \\ 
  \hline
Shift=0 & 0.000 & 0.000 & 0.001 & 0.002 & 0.005 & 0.020 & 0.112 \\ 
  Shift=1 & 0.000 & 0.000 & 0.000 & 0.001 & 0.002 & 0.005 & 0.019 \\ 
  Shift=2 & 0.000 & 0.000 & 0.000 & 0.001 & 0.001 & 0.003 & 0.006 \\ 
  Shift=3 & 0.018 & 0.007 & 0.002 & 0.000 & 0.002 & 0.003 & 0.004 \\ 
  Shift=4 & 0.180 & 0.100 & 0.040 & 0.010 & 0.002 & 0.005 & 0.005 \\ 
  Shift=5 & 0.579 & 0.432 & 0.261 & 0.108 & 0.027 & 0.008 & 0.009 \\ 
   \hline
\end{tabular}
\caption{HAC adjusted p-values of regressions of BIP (M-SSA) component optimized for forecast horizons $h=0,...,6$  (the columns) on HP-BIP shifted forward by shift=0,...,5 (the rows). Out-of-sample span starting in Jan-2007, without pandemic.} 
\label{p_val_hpbip_wc}
\end{table}Our choice—focusing on the single M-SSA BIP component—emphasizes simplicity and interpretability, as future BIP and HP-BIP are connected through their shared low-frequency component. Additionally, the M-SSA BIP component is the most important predictor in the regression of the M-SSA outputs on BIP.  Finally, revisions due to quarterly updates of the regression equations are minimized by this predictor, and the estimates are more stable and robust over time\footnote{Other combinations can be experimented with using the M-SSA package. Regression weights of more complex designs, involving multiple components, tend to be more difficult to interpret due to the occurrence of negative weights and potential sign changes of the regressors over time, which may indicate overfitting or non-stationarity.} (see Section \eqref{revisions}). Note that information from the other  indicators is incorporated into the M-SSA BIP component through the multivariate filter design.\\

@@

% latex table generated in R 4.2.2 by xtable 1.8-4 package
% Thu May  8 07:34:09 2025
\begin{table}[ht]
\centering
\begin{tabular}{rrrrrrrr}
  \hline
 & h=0 & h=1 & h=2 & h=3 & h=4 & h=5 & h=6 \\ 
  \hline
Shift=0 & 0.084 & 0.047 & 0.017 & 0.005 & 0.002 & 0.001 & 0.038 \\ 
  Shift=1 & 0.230 & 0.160 & 0.094 & 0.032 & 0.005 & 0.001 & 0.004 \\ 
  Shift=2 & 0.801 & 0.666 & 0.414 & 0.173 & 0.026 & 0.005 & 0.006 \\ 
  Shift=3 & 0.876 & 0.961 & 0.917 & 0.683 & 0.248 & 0.068 & 0.039 \\ 
  Shift=4 & 0.370 & 0.819 & 0.990 & 0.953 & 0.533 & 0.117 & 0.030 \\ 
  Shift=5 & 0.353 & 0.802 & 0.983 & 0.943 & 0.571 & 0.239 & 0.043 \\ 
   \hline
\end{tabular}
\caption{HAC adjusted p-values of regressions of M-SSA component predictors optimized for forecast horizons $h=0,...,6$  (the columns) on BIP shifted forward by shift=0,...,5 (the rows). Out-of-sample span starting in Jan-2007, without pandemic.} 
\label{p_val_wc}
\end{table}% latex table generated in R 4.2.2 by xtable 1.8-4 package
% Thu May  8 07:34:09 2025
\begin{table}[ht]
\centering
\begin{tabular}{rrrrrrrr}
  \hline
 & h=0 & h=1 & h=2 & h=3 & h=4 & h=5 & h=6 \\ 
  \hline
Shift=0 & 0.978 & 0.942 & 0.893 & 0.855 & 0.860 & 0.906 & 0.962 \\ 
  Shift=1 & 1.082 & 1.060 & 1.005 & 0.926 & 0.877 & 0.880 & 0.914 \\ 
  Shift=2 & 1.083 & 1.090 & 1.077 & 1.019 & 0.952 & 0.918 & 0.919 \\ 
  Shift=3 & 1.034 & 1.059 & 1.081 & 1.063 & 1.010 & 0.970 & 0.954 \\ 
  Shift=4 & 0.995 & 1.029 & 1.069 & 1.077 & 1.035 & 0.986 & 0.953 \\ 
  Shift=5 & 0.993 & 1.030 & 1.075 & 1.093 & 1.062 & 1.011 & 0.967 \\ 
   \hline
\end{tabular}
\caption{rRMSEs of M-SSA components predictor benchmarked against the expanding mean of BIP. Out-of-sample span starting in Jan-2007, without pandemic.} 
\label{rRMSE_mSSA_comp_mean_wc}
\end{table}

>>>>

% latex table generated in R 4.4.1 by xtable 1.8-4 package
% Thu May  8 14:00:18 2025
\begin{table}[ht]
\centering
\begin{tabular}{rrrrrr}
  \hline
 & GDP & ip & ifo\_c & ESI & spr\_10y\_3m \\ 
  \hline
2024-08-01 &  & 93.200 & 84.900 & 90.900 & -1.400 \\ 
  2024-09-01 &  & 91.200 & 83.700 & 89.800 & -1.400 \\ 
  2024-10-01 & 900.761 & 90.800 & 84.100 & 90.700 & -1.200 \\ 
  2024-11-01 &  & 92.200 & 83.800 & 89.300 & -0.900 \\ 
  2024-12-01 &  &  & 82.800 & 86.900 & -0.800 \\ 
  2025-01-01 &  &  & 82.500 & 88.100 & -0.200 \\ 
   \hline
\end{tabular}
\caption{Ragged end.  } 
\label{f_stat}
\end{table}% latex table generated in R 4.4.1 by xtable 1.8-4 package
% Thu May  8 14:00:18 2025
\begin{table}[ht]
\centering
\begin{tabular}{rrrrrr}
  \hline
 & GDP & ip & ifo\_c & ESI & spr\_10y\_3m \\ 
  \hline
2024-05-01 &  & 93.400 & 84.900 & 90.900 & -1.400 \\ 
  2024-06-01 &  & 90.700 & 83.700 & 89.800 & -1.400 \\ 
  2024-07-01 & 902.571 & 93.200 & 84.100 & 90.700 & -1.200 \\ 
  2024-08-01 &  & 91.200 & 83.800 & 89.300 & -0.900 \\ 
  2024-09-01 &  & 90.800 & 82.800 & 86.900 & -0.800 \\ 
  2024-10-01 & 900.761 & 92.200 & 82.500 & 88.100 & -0.200 \\ 
   \hline
\end{tabular}
\caption{Data aligned at sample end.} 
\label{f_stat}
\end{table}

@@

So-called direct forecasts are obtained by regressing the relevant indicators on forward-shifted BIP, accounting for the additional publication lag. For illustration of the concept, we here rely on all indicators and full sample information, excluding the pandemic. In-sample forecasts and the shifted BIP, which serves as the target, are presented in Figs. \ref{direct_wc} (full dataset without the pandemic) and \ref{direct_wc_financial_crisis} (financial crisis) for horizons up to three quarters ahead. 
\begin{figure}[H]\begin{center}\includegraphics[height=5in, width=6in]{./Figures/direct_wc_all.pdf}\caption{Direct forecasts (red lines) based on all indicators and full sample information, without the pandemic: GDP (black line) shifted by zero one, two and three quarters.\label{direct_wc}}\end{center}\end{figure}\begin{figure}[H]\begin{center}\includegraphics[height=5in, width=6in]{./Figures/direct_wc_financial_crisis.pdf}\caption{Direct forecast  across the financial crisis:  regression based on all indicators and full sample information.\label{direct_wc_financial_crisis}}\end{center}\end{figure}The results indicate that as $h$ increases, the peaks and dips of the direct forecasts become increasingly right-shifted (delayed) relative to the target. Additionally, the F-statistics precipitously decline from $F=3.57$, for $h=2$, to $F=0.46$, for $h=3$, see Table \eqref{f_stat}, suggesting that the predictor becomes statistically insignificant for larger forecast horizons exceeding two quarters (link to Heinisch and Scheufele).

@@

As shown, the performance of direct forecasts sharply declines for larger forecast horizons. We conjecture that erratic short-term fluctuations -—unpredictable high-frequency noise -— overlay the data, thereby obscuring the effective 'signal' and making a direct regression more susceptible to overfitting. We therefore envisage to highlight the signal by applying a filter to dampen high-frequency noise. 
The so-called Hodrick-Prescott filter is a classic tool used in business cycle analysis: the filter is specified by a single smoothing parameter $\lambda$ and the value $\lambda=1600$ is specifically recommended for quarterly data. However, Phillips and Jin (2021) suggest that the HP(1600) filter removes relevant information due to excessive smoothing.  In our context, this oversmoothing issue would be further exacerbated when considering forecast horizons shorter than a year -—an interval inconsistent with the mean duration of up to several years of business cycles, as highlighted by the HP(1600). Consequently, we here select a more adaptive HP(160) 'target' filter. 
While this choice may appear somewhat arbitrary, an analysis of the filter in the frequency domain confirms a more suitable profile for the corresponding amplitude function, assigning greater weight to yearly components -—relevant in our prediction framework-— than the classic quarterly HP filter. This approach also addresses and controls for undesirable high-frequency noise. See Fig. \ref{hp_160}, which presents both two-sided and (classic) one-sided filters (top panels) along with the associated amplitude functions (bottom panels). We now posit the HP(160) filter as a means to achieve better forecasting performance at larger horizons, noting that a comprehensive technical analysis of the effect of $\lambda$ on the resulting BIP predictor can be conducted using the M-SSA package. 
\begin{figure}[H]\begin{center}\includegraphics[height=5in, width=6in]{./Figures/hp_160.pdf}\caption{Two-sided HP-160) (top left) and one-sided (concurrent) HP-C filters (top right) with corresponding amplitude functions (bottom).\label{hp_160}}\end{center}\end{figure}
\subsection{Direct HP Forecast}\label{hpdf}

Extending the `classic' direct forecast design, we here consider filtered indicators, based on the one-sided HP(160), denoted as HP-C: the filtered indicators are explanatory variables for the regressions on forward-shifted BIP\footnote{Katja and Simon: for each forecast horizon $h=0,1,2,3$ I'm using corresponding forecasts of the real-time filter output, assuming the data to be white noise (assuming an ARMA model does not lead to significantly different results). I think we might ignore these technical details in the paper.}. As in the previous section, we rely on all indicators as well as the full sample estimate for illustration, see Figs.\eqref{direct_hp_forecasts} (all observations with exclusion of the pandemic) and \eqref{direct_hp_forecasts_financial_crisis} (financial crisis). A comparison with the classic direct forecasts in Fig.\eqref{direct_wc_financial_crisis} indicates that the new predictors are slightly less retarded (`faster'). 
\begin{figure}[H]\begin{center}\includegraphics[height=5in, width=6in]{./Figures/direct_hp_forecasts.pdf}\caption{Direct HP forecasts: entire data set with exclusion of the pandemic.\label{direct_hp_forecasts}}\end{center}\end{figure}\begin{figure}[H]\begin{center}\includegraphics[height=5in, width=6in]{./Figures/direct_hp_forecasts_financial_crisis.pdf}\caption{Shifted GDP (black), direct HP forecast (blue) and classic direct forecast (red) over the curse of the financial crisis:  regression based on all indicators and full sample information (without pandemic).\label{direct_hp_forecasts_financial_crisis}}\end{center}\end{figure}Moreover, the F-statistics are larger than for the direct forecasts, see Table \eqref{f_stat}, but the predictor remains statistically insignificant at forecast horizons exceeding two quarters (even though the evaluation is conducted in-sample and based on all available indicators). 
% latex table generated in R 4.4.1 by xtable 1.8-4 package
% Thu May  8 14:00:18 2025
\begin{table}[ht]
\centering
\begin{tabular}{rrrrr}
  \hline
 & h=0 & h=1 & h=2 & h=3 \\ 
  \hline
Direct forecast & 7.188 & 5.285 & 3.574 & 0.460 \\ 
  Direct HP forecast & 21.886 & 8.415 & 4.810 & 1.834 \\ 
   \hline
\end{tabular}
\caption{F-statistics of direct forecasts and direct HP forecasts for forecast horizons $h=0,1,2,3$: fully in-sample.  } 
\label{f_stat}
\end{table}We now propose a more refined causal filter design aiming for statistical significance of the resulting predictor at longer forecast horizons, out-of-sample.

@@

Figure \ref{bk_gammak} displays the coefficients of the multivariate nowcasts, where increased smoothing achieved through the M-SSA method manifests as a distinct pattern in the filter weights, notably influencing the rate of decay across increasing lags. The resulting filtered series, presented in Figure \ref{mssa_msse_now}, are standardized to enable direct comparisons. The M-SSA nowcast adheres to the specified constraint, exhibiting approximately one-third fewer zero-crossings than the M-MSE, confirming the effectiveness of the smoothness control imposed during the design. This is further corroborated by Table \ref{corhtnow}, which reports empirical holding times and target correlations. Increased smoothness achieved by the M-SSA comes at the cost of a slight reduction in tracking accuracy, as indicated by a marginally lower target correlation. This highlights the inherent trade-off—often referred to as the prediction dilemma—between smoothness (zero-crossing rate) and predictive accuracy (target correlation) within the M-SSA criterion.
\begin{figure}[H]\begin{center}\includegraphics[height=3in, width=5in]{./Figures/bk_gammak.pdf}\caption{Multivariate M-MSE (left) and M-SSA  (right) nowcast filters.\label{bk_gammak}}\end{center}\end{figure}\begin{figure}[H]\begin{center}\includegraphics[height=4in, width=5.5in]{./Figures/mssa_msse_now.pdf}\caption{Two-sided HP(160) applied to GDP (target, black) and nowcasts:  M-SSA (blue) and  M-MSE (green). The dashed vertical line delimits in- and out-of-sample spans. The two-sided filter does not extend to the sample end.\label{mssa_msse_now}}\end{center}\end{figure}% latex table generated in R 4.4.1 by xtable 1.8-4 package
% Thu May  8 14:00:18 2025
\begin{table}[ht]
\centering
\begin{tabular}{rrr}
  \hline
 & M-SSA & M-MSE \\ 
  \hline
Target correlation & 0.68 & 0.69 \\ 
  HT & 6.50 & 3.71 \\ 
   \hline
\end{tabular}
\caption{Sample target correlations and HTs of M-SSA and M-MSE nowcasts.  } 
\label{corhtnow}
\end{table}The results validate that the smoothing constraint incorporated into the multivariate extension of Criterion \eqref{critssa} effectively manages the zero-crossings of the predictor: sample estimates of the holding times (HTs) align with the expected values derived from Equation \eqref{ht}\footnote{Effective convergence of sample estimates to their theoretical expected values —assuming the true data-generating process is known— can be validated using the M-SSA package. This validation is typically performed through analyses of very long samples of simulated (multivariate) data, ensuring the consistency of the estimation procedures under idealized conditions.}. An increase of $50\%$ in the HT relative to the M-MSE predictor has only a modest effect on the target correlation, suggesting that this tradeoff could be beneficial in our application. Going forward, we now omit the M-MSE, as it is a special case of the more general M-SSA framework.

@@

M-SSA predictors are compared to the traditional univariate HP-C in Figs. \eqref{multivar_vs_univar} and \eqref{mssa_hpc_financial_crisis}. In the first figure, we compare both filters across the periods of the financial crisis (top panels) and the COVID-19 pandemic (bottom panels), for forecast horizons $h=0,...,6$. The nowcast from the previous section is included for reference. In the second figure we compare both filters to the forward-shifted BIP for shifts larger than two quarters. All series are standardized to facilitate visual comparison. The primary distinction of the M-SSA approach is the progressive left-shift of the predictor as a function of $h$; this shift becomes more pronounced with increasing $h$ and appears to be consistent across all levels, including the peaks and troughs of the series. Conversely, the positions of peaks and troughs in the HP-C series appear to remain largely unaffected by variations in $h$. The more pronounced and systematic left-shift observed in the M-SSA forecasts can be partly attributed to the increasing weight assigned to the additional indicators leading BIP, which dominate the BIP component at larger forecast horizons. %In comparison, the effect of phase shifts—arising from the cyclical nature of the HP-filter and manifesting in the HP-C forecasts (shown in the left panels)—seems less systematic. 
\begin{figure}[H]\begin{center}\includegraphics[height=3in, width=5.5in]{./Figures/bk_h.pdf}\caption{Multivariate M-SSA: $h=0$ (nowcast, left) and $h=4$ (one year forecast, right).\label{bk_h}}\end{center}\end{figure}\begin{figure}[H]\begin{center}\includegraphics[height=4in, width=5.5in]{./Figures/multivar_vs_univar.pdf}\caption{Now- and forecasts of HP-GDP for $h=0,...,6$: HP-C (top) vs. M-SSA (bottom) across the financial crisis (left) and the pandemic (right).\label{multivar_vs_univar}}\end{center}\end{figure}
\begin{figure}[H]\begin{center}\includegraphics[height=4in, width=5.5in]{./Figures/mssa_hpc_financial_crisis.pdf}\caption{Forward-shifted GDP (black), M-SSA (blue) and HP-C (red) for $h=3,4,5,6$ across the financial crisis. All series are standardized for better visual inspection.\label{mssa_hpc_financial_crisis}}\end{center}\end{figure}

@@

There are various combinations of M-SSA components for tracking BIP. For illustration, we here select the single BIP component resulting from the M-SSA filter tracking HP-BIP. Table \eqref{p_val_hpbip_wc} reports HAC-adjusted p-values from regressions of the out-of-sample predictor on HP-BIP, indicating a strong link between the predictor and the target, out-of-sample. 
% latex table generated in R 4.4.1 by xtable 1.8-4 package
% Thu May  8 14:00:20 2025
\begin{table}[ht]
\centering
\begin{tabular}{rrrrrrrr}
  \hline
 & h=0 & h=1 & h=2 & h=3 & h=4 & h=5 & h=6 \\ 
  \hline
Shift=0 & 0.000 & 0.000 & 0.001 & 0.002 & 0.005 & 0.020 & 0.112 \\ 
  Shift=1 & 0.000 & 0.000 & 0.000 & 0.001 & 0.002 & 0.005 & 0.019 \\ 
  Shift=2 & 0.000 & 0.000 & 0.000 & 0.001 & 0.001 & 0.003 & 0.006 \\ 
  Shift=3 & 0.018 & 0.007 & 0.002 & 0.000 & 0.002 & 0.003 & 0.004 \\ 
  Shift=4 & 0.180 & 0.100 & 0.040 & 0.010 & 0.002 & 0.005 & 0.005 \\ 
  Shift=5 & 0.579 & 0.432 & 0.261 & 0.108 & 0.027 & 0.008 & 0.009 \\ 
   \hline
\end{tabular}
\caption{HAC adjusted p-values of regressions of BIP (M-SSA) component optimized for forecast horizons $h=0,...,6$  (the columns) on HP-BIP shifted forward by shift=0,...,5 (the rows). Out-of-sample span starting in Jan-2007, without pandemic.} 
\label{p_val_hpbip_wc}
\end{table}Our choice—focusing on the single M-SSA BIP component—emphasizes simplicity and interpretability, as future BIP and HP-BIP are connected through their shared low-frequency component. Additionally, the M-SSA BIP component is the most important predictor in the regression of the M-SSA outputs on BIP.  Finally, revisions due to quarterly updates of the regression equations are minimized by this predictor, and the estimates are more stable and robust over time\footnote{Other combinations can be experimented with using the M-SSA package. Regression weights of more complex designs, involving multiple components, tend to be more difficult to interpret due to the occurrence of negative weights and potential sign changes of the regressors over time, which may indicate overfitting or non-stationarity.} (see Section \eqref{revisions}). Note that information from the other  indicators is incorporated into the M-SSA BIP component through the multivariate filter design.\\

@@

% latex table generated in R 4.4.1 by xtable 1.8-4 package
% Thu May  8 14:00:20 2025
\begin{table}[ht]
\centering
\begin{tabular}{rrrrrrrr}
  \hline
 & h=0 & h=1 & h=2 & h=3 & h=4 & h=5 & h=6 \\ 
  \hline
Shift=0 & 0.084 & 0.047 & 0.017 & 0.005 & 0.002 & 0.001 & 0.038 \\ 
  Shift=1 & 0.230 & 0.160 & 0.094 & 0.032 & 0.005 & 0.001 & 0.004 \\ 
  Shift=2 & 0.801 & 0.666 & 0.414 & 0.173 & 0.026 & 0.005 & 0.006 \\ 
  Shift=3 & 0.876 & 0.961 & 0.917 & 0.683 & 0.248 & 0.068 & 0.039 \\ 
  Shift=4 & 0.370 & 0.819 & 0.990 & 0.953 & 0.533 & 0.117 & 0.030 \\ 
  Shift=5 & 0.353 & 0.802 & 0.983 & 0.943 & 0.571 & 0.239 & 0.043 \\ 
   \hline
\end{tabular}
\caption{HAC adjusted p-values of regressions of M-SSA component predictors optimized for forecast horizons $h=0,...,6$  (the columns) on BIP shifted forward by shift=0,...,5 (the rows). Out-of-sample span starting in Jan-2007, without pandemic.} 
\label{p_val_wc}
\end{table}% latex table generated in R 4.4.1 by xtable 1.8-4 package
% Thu May  8 14:00:20 2025
\begin{table}[ht]
\centering
\begin{tabular}{rrrrrrrr}
  \hline
 & h=0 & h=1 & h=2 & h=3 & h=4 & h=5 & h=6 \\ 
  \hline
Shift=0 & 0.978 & 0.942 & 0.893 & 0.855 & 0.860 & 0.906 & 0.962 \\ 
  Shift=1 & 1.082 & 1.060 & 1.005 & 0.926 & 0.877 & 0.880 & 0.914 \\ 
  Shift=2 & 1.083 & 1.090 & 1.077 & 1.019 & 0.952 & 0.918 & 0.919 \\ 
  Shift=3 & 1.034 & 1.059 & 1.081 & 1.063 & 1.010 & 0.970 & 0.954 \\ 
  Shift=4 & 0.995 & 1.029 & 1.069 & 1.077 & 1.035 & 0.986 & 0.953 \\ 
  Shift=5 & 0.993 & 1.030 & 1.075 & 1.093 & 1.062 & 1.011 & 0.967 \\ 
   \hline
\end{tabular}
\caption{rRMSEs of M-SSA components predictor benchmarked against the expanding mean of BIP. Out-of-sample span starting in Jan-2007, without pandemic.} 
\label{rRMSE_mSSA_comp_mean_wc}
\end{table}
